---
title: "CapDAP_analysis"
author: "Marliese"
date: "11/14/2022"
output: html_document
---

###Introduction:


###Analysis:

First lets set things up
```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(readr)
```


##Analysis 1:

1. pull in data

```{r}
WsNumSp <- read.csv(here("Data", "Processed Data", "Ws_Num_Sp.csv"))
```

2. plot expected relationships

```{r}
ggplot(WsNumSp, aes(x = Watershed, y=length.unique.Species..))+
     geom_point()+
  xlab("Watershed")+
  ylab("Number of Species Present")
```


3. guess expected relationships (hypothesis)

I think there will not be a significant difference in the number of species

4. decide on statistical model and test and explain why

This will be a chi squared test because both Watershed and Species are catagorical nominal variables.

5. check model assumptions

6. run stat model

7. interpret model results (were expected relationships supported?)

8. replot data to show key relationships


##Analysis 2:

1. pull in data

2. plot expected relationships

3. guess expected relationships (hypothesis)

4. decide on statistical model and test and explain why

5. check model assumptions

6. run stat model

7. interpret model results (were expected relationships supported?)

8. replot data to show key relationships

##Analysis 3:

1. pull in data

2. plot expected relationships

3. guess expected relationships (hypothesis)

4. decide on statistical model and test and explain why

5. check model assumptions

6. run stat model

7. interpret model results (were expected relationships supported?)

8. replot data to show key relationships

##Analysis 4:

1. pull in data

2. plot expected relationships

3. guess expected relationships (hypothesis)

4. decide on statistical model and test and explain why

5. check model assumptions

6. run stat model

7. interpret model results (were expected relationships supported?)

8. replot data to show key relationships




###Biological Summary:


###Challenges:



