---
title: "Exploratory_Stuff"
author: "Marliese"
date: "11/17/2022"
output: html_document
---

First lets set things up
```{r}
rm(list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(readr)
```

Now let's pull in the data
```{r}
DF <- read.csv(here("Data", "CapDAP_data.csv"))
```


###Part1: number of species per watershed (species richness)

lets make the first dataframe for analysis 1



1. Make Watershed and species a factor variable
```{r}
DF$Watershed<-as.factor(DF$Watershed)
DF$Species<-as.factor(DF$Species)
```


2. look at the levels
```{r}
levels(DF$Watershed)
levels(DF$Species)
```

Initially, SpB and SPB look to me to be the same, and after looking at the metadata I am confident they are. So I am going to fix that really quick before moving on.

3. fix levels of watershed
```{r}
fix <- which(DF$Watershed == "SPB")
DF$Watershed[fix]<-"SpB"
```

R still thinks there are 9 levels. Lets relevel them by making Watershed a factor variable again.

4. relevel
```{r}
DF$Watershed <-factor(DF$Watershed)
levels(DF$Watershed)
```

5. plot expected relationship
```{r}
ggplot(DF, aes(x = Watershed,  y = Species))+
     geom_point()+
     theme_bw()
```
As this plot shows, not all species are found in each watershed.

6. make a new dataframe to show just number of species found in each watershed
```{r}
newdata<-DF%>%
group_by(Watershed)%>%
  summarise(
length(unique(Species)))
```


7. Copy this new dataframe and put it in data folder ***MAKE SURE THIS IS CORRECT before running
```{r}
write.csv(newdata, here("Data","Processed Data","Ws_Num_Sp.csv"), row.names = FALSE) #row.names=FALSE means please dont make X column where it has numbered the rows
```



###Part2: 2-way anova avg weight of spp by watershed

lets make the first dataframe for analysis 2. First start over and clear environment and re-pull in the data.

1. look at variables and see if need to change them at all
```{r}

DF$Watershed<-as.factor(DF$Watershed)
DF$Weight<-as.numeric(DF$Weight)
```


2. look at the levels
```{r}
levels(DF$Watershed)
levels(DF$Weight) #not actually something you can look at but I put it in just to be thorough
```

Initially, SpB and SPB look to me to be the same, and after looking at the metadata I am confident they are. So I am going to fix that really quick before moving on.

3. fix levels of watershed
```{r}
fix <- which(DF$Watershed == "SPB")
DF$Watershed[fix]<-"SpB"
```

R still thinks there are 9 levels. Lets relevel them by making Watershed a factor variable again.

4. relevel
```{r}
DF$Watershed <-factor(DF$Watershed)
levels(DF$Watershed)
```

5. plot expected relationship
```{r}
ggplot(DF, aes(Watershed, Weight))+
     geom_boxplot()+
     geom_point(size = 2, alpha = 0.5, position = "jitter")+
        facet_wrap(~Species)
```

As this plot shows, there are a lot of other things going on. To get rid of some of this noise I am going to subset to look at individuals that are Adults, and only look at a few species.


6. make age and species factor variables bc I will be using them
```{r}
DF$Species<-as.factor(DF$Species)
DF$Age<-as.factor(DF$Age)
```

7. look at the levels of Age (I know the levels of Species are good bc I looked at them in part 1)
```{r}
levels(DF$Age)
```

Okay, so I looked at the metadata and based on that I am going to fix some things

8. fix levels
```{r}
no <- which(DF$Age == "JUV"| DF$Age == "J")
DF$Age[no]<-"Juv"

nah <- which(DF$Age == "?")
DF$Age[nah]<-""
```

9. relevel
```{r}
DF$Age <-factor(DF$Age)
levels(DF$Age)
```

That looks better. There is still a weird "SJ" (which I am assuming to mean sub-juvenile but the metadata doesn't actually clarify). If I was going to look at Weight in relation to age, which would be interesting, I would omit "SJ" and "", however because I am just going to be looking at Adults I am going to leave it.


10. make a new dataframe to show just number of species found in each watershed

I am going to do it in levels

10.a) subset just adults
```{r}
adults<-DF%>%
        filter(Age=="A")
```

look at the spread now
```{r}
ggplot(adults, aes(Watershed, Weight))+
     geom_boxplot()+
     geom_point(size = 2, alpha = 0.5, position = "jitter")+
        facet_wrap(~Species)
```
Looks good but we can make it better.


10.b) subset to specific species
```{r}
OldSpp<-adults%>%
  filter(Species=="SIHI"|Species=="MIOC")
```

look at spread
```{r}
ggplot(OldSpp, aes(Watershed, Weight))+
     geom_boxplot()+
     geom_point(size = 2, alpha = 0.5, position = "jitter")+
        facet_wrap(~Species)
```
Okay, this looks better. I am going to keep these two species. But first I need to separate them out into different dataframes.

```{r}
OldSIHI<-adults%>%
  filter(Species=="SIHI")

OldMIOC<-adults%>%
  filter(Species=="MIOC")
```

10.c) omit na values so I can proceed
```{r}
OldSIHI <- OldSIHI[!(is.na(OldSIHI$Weight)), ]
OldMIOC <- OldMIOC[!(is.na(OldMIOC$Weight)), ]
```


10.d) make new dataframes
```{r}
SIHI_Ws_AvgW<-OldSIHI%>%
  group_by(Watershed)%>%
  summarise(
    avg_Weight=mean(Weight))
```


```{r}
MIOC_Ws_AvgW<-OldMIOC%>%
  group_by(Watershed)%>%
  summarise(
    avg_Weight=mean(Weight))
```


11. Copy this new dataframe and put it in data folder ***MAKE SURE THIS IS CORRECT before running
```{r}
write.csv(SIHI_Ws_AvgW, here("Data","Processed Data","SIHI_Ws_AvgW.csv"), row.names = FALSE) #row.names=FALSE means please dont make X column where it has numbered the rows

write.csv(MIOC_Ws_AvgW, here("Data","Processed Data","MIOC_Ws_AvgW.csv"), row.names = FALSE) #row.names=FALSE means please dont make X column where it has numbered the rows
```




###Part 3: species composition and watershed

lets make the dataframe for analysis 3. First start over and clear environment and re-pull in the data.

1. Make Watershed and species a factor variable
```{r}
DF$Watershed<-as.factor(DF$Watershed)
DF$Species<-as.factor(DF$Species)
DF$Transect<-as.factor(DF$Transect)
```


2. look at the levels
```{r}
levels(DF$Watershed)
levels(DF$Species)
```

Initially, SpB and SPB look to me to be the same, and after looking at the metadata I am confident they are. So I am going to fix that really quick before moving on.

3. fix levels of watershed
```{r}
fix <- which(DF$Watershed == "SPB")
DF$Watershed[fix]<-"SpB"
```

R still thinks there are 9 levels. Lets relevel them by making Watershed a factor variable again.

4. relevel
```{r}
DF$Watershed <-factor(DF$Watershed)
levels(DF$Watershed)
```

5. plot expected relationship
```{r}
ggplot(DF, aes(x = Watershed,  y = Species))+
     geom_point()+
     theme_bw()
```
***note this is not the expected relationship. but move on now



6. subset data so only have dataframe w/ date, watershed, transect, and species. Look at that dataframa and then create another dataframe, and add a column w the number of individuals per species per transect in each watershed.
```{r}
DWTS<-DF%>%
  subset(select=c(Date,Watershed,Transect,Species))

try<-DF%>%
  subset(select=c(Date,Watershed,Species))%>%
  group_by(Watershed,Species)%>%
  summarise(
    length(Species))
```
note: did this and included Transect in the subsets and in the group_by for df 'try', but after a lot of work and troubleshooting, I went back and took it out because then the Shannon idex was giving me the diversity for each transect not each watershed.


7. make length(Species) a numeric variable bc vegan gives a lot of errors if things aren't
```{r}
try$`length(Species)`<-as.numeric(try$`length(Species)`)
```



8. try to pivot_wider just the species column so each species is a column header and the value in the cells are the number of inidivids
```{r}
WTS_wide<-try%>%
  pivot_wider(names_from = "Species", values_from = "length(Species)") 
```
wait that like actually did what I wanted it to!! yay!


9. use vegan package to get Shannon index and Simpson index

9.a) first load vegan
```{r}
library(vegan)
library(ggvegan)
```
note: go to this page to learn abt vegan https://cran.r-project.org/web/packages/vegan/vignettes/diversity-vegan.pdf 
or this one
https://rpubs.com/an-bui/vegan-cheat-sheet


9.b) lets try some things to get the shannon diversity index

okay make NA = 0 in WTS_wide
```{r}
WTS_wide[is.na(WTS_wide)] = 0
```


run shannon diversity index on that
```{r}
Sh_div<-diversity(WTS_wide[-1], index="shannon")
```
note: when I run the following code it gives me a different number than the one it gave for the code above this note.
```{r}
#diversity(WTS_wide$MIOC, index = "shannon",
         # MARGIN = 1, base = exp(1))
```



10. add that column to dataframe!!!
```{r}
WTS_wide$Sh_div <-Sh_div
```
yes yes yes!


11. Copy this new dataframe and put it in processed data folder ***MAKE SURE THIS IS CORRECT before running
```{r}
write.csv(WTS_wide, here("Data","Processed Data","Ws_Sp_Shannon.csv"), row.names = FALSE) #row.names=FALSE means please dont make X column where it has numbered the rows
```




